---
title: "IODS course project"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    fig_caption: true
    fig_width: 8
    fig_height: 6
    code_folding: none
author: 
  name: "Juho Pirhonen"
  email: juho.pirhonen@helsinki.fi
date: "18/12/2017"
css: styles.css
---
<script src=showCodePanel.js type="text/javascript" character="UTF-8"></script>

# Load data
```{R}
#Hidden
source("helper_functions.R")
library(knitr)
library(kableExtra)
library(ggplot2)
library(GGally)
library(tidyr)
library(stringr)
library(dplyr, warn.conflicts = F)
df <- read.csv("data/human.csv") %>% tbl_df()
```
This dataset contains `r nrow(df)` obseravtions and `r ncol(df)-1` variables exploring dependancy between proxy variables of human developement and gender (in)equality by country. The data originates from http://hdr.undp.org/en/data. Details of data-wrangling can be found [here](https://github.com/Juhous/create_human.R).

## Study variables

| Variable | Description |
|------|---------|
| GNIC | Gross National Income per capita |
| lifeExp | Life expectancy | 
| eduExp | Expected years of education |
| matMort | Maternal mortality ratio |
| adolBirthRate | Adolescent birth rate |
| reprParl | Percetange of female representatives in parliament |
| edu2F | Proportion of females with at least secondary education |
| edu2M | Proportion of males with at least secondary education |
| labF | Proportion of females in the labour force |
| labM | Proportion of males in the labour force |
| eduRatio | ratio edu2F and edu2M |
| labRatio | ratio edu2F and edu2M |


# Exploratory Data Analysis
## Numercial overview
```{r}
#Hidden
df %>% select(-country) %>% summaryKable() %>% 
  kable("html", align = "rrr", caption = "Study variable summary") %>%
  kable_styling(bootstrap_options = c("hover", "condensed")) %>% 
  scroll_box(width = "100%")
```

## Graphical overview
```{R}
#Hidden
df %>% gather(key="var", value = "value", HDI:labRatio) %>%
  mutate(var=ordered(var, names(df[2:17]))) %>%  #To retain order when facetting
  ggplot() +
  geom_density(aes(value)) + 
  facet_wrap(~var, scales="free") +
  labs(title = "Study variable distribution", 
       y = "Density", x="Value") +
  theme(axis.text.x = element_text(angle = 90))
```

Most variables are distributed normally, but GNIC and matMort might benefit from logarithmic conversion. 

```{r}
#Hidden
# Save a plot of non-normally distributed variables
p1 <- df %>% 
  gather(key="var", value = "value", c(GNIC, matMort)) %>%
  mutate(var=ordered(var, names(df[2:17]))) %>%  #To retain order when facetting
  ggplot() +
  geom_density(aes(value)) + 
  facet_wrap(~var, scales="free") +
  labs(title = "Before log-conv", 
       y = "Density", x="Value") 

#Log conversion for non-normal values
df %<>% mutate_at(vars(GNIC, matMort, adolBirthRate), funs(log10)) %>% 
  rename_at(vars(GNIC, matMort, adolBirthRate), 
            funs(paste(., "log", sep="_")))

# Store a plot of new values
p2 <- df %>% 
  gather(key="var", value = "value", c(GNIC_log, matMort_log)) %>%
  mutate(var=ordered(var, names(df[2:17]))) %>%  #To retain order when facetting
  ggplot() +
  geom_density(aes(value)) + 
  facet_wrap(~var, scales="free") +
  labs(title = "After log-conv", 
       y = "Density", x = "Value") 

#Show plots
multiplot(p1,p2, cols = 1)

write.csv(df, "data/df.csv")
```

Now the variables are distributed more normally. 
