# Linear discriminant analysis
```{r, echo = F}
#Hidden
source("helper_functions.R")
library(magrittr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(GGally)
library(tidyr, warn.conflicts = F)
library(stringr)
library(MASS, warn.conflicts = F) #NOTE
library(dplyr, warn.conflicts = F)
#note
df <- read.csv("data/df_inliers.csv") %>% select(-c(GII, eduRatio, eduRatioT))
row.names(df) <- df$country
df %<>% select(-country)
```
Next I wanted to explore LDA, a form of dimensionality reduction. 

## LDA
First, the dependent variable, HDI, is divided to 4 equally sized categories. All other variables are scaled to have 
```{R}
#Hidden
# Explanatory variables are scaled
df %<>% mutate_all(funs(scale))
# sapply(df, sd)

# HDI is divided into quartiles
df$HDI <- ntile(df$HDI, 4) %>% 
  factor(labels = c("low", "med-lo", "med-hi", "high"))

# The data is sampled into training and testing sets
sample_ind <- sample(nrow(df),  size = nrow(df) * 0.8)
train <- df[sample_ind,]
test <-  df[-sample_ind,]
#data.frame(dim(train),dim(test))

#LDA
lda.fit <- lda(HDI ~ ., data = train)
lda.fit
```

Almost all variation in the data is explained just by maternal mortality and education inequality. 

## Visualising the mode
```{r, out.width = "100%"}
#Hidden
library(plotly, warn.conflicts = F)

# Plot the lda results
points <- data.frame(HDI = train$HDI,
                     lda = predict(lda.fit)$x)
levels(points$HDI) %<>%  str_to_title()

arrows <- coef(lda.fit) %>% 
  data.frame(., label = rownames(.)) %>% arrange(desc(abs(LD1))) %>% 
  mutate(LD1 = LD1*2.5, LD2 = LD2*2.5, LD3 = LD3*2.5, pos = 1) %>% 
  rbind(., mutate(., LD1=0, LD2=0, LD3=0, pos =0)) 


p1 <- plot_ly(arrows, x = ~LD1, y = ~LD2, z = ~LD3, 
  type = "scatter3d" , color = ~label, colors = rep(rgb(0, 0, 0), 13),
  opacity = .5, mode = "lines", hoverinfo = "name", showlegend = FALSE, 
  line = list(width = 5))

p2 <- plot_ly(points, x = ~lda.LD1, y = ~lda.LD2, z = ~lda.LD3, 
    type = "scatter3d" , color = ~HDI, opacity = .75, hoverinfo = "none",
    mode = "markers", marker = list(size = 3, width = 2)) %>% 
  layout(title = "PCA",
       scene = list(xaxis = list(title = "LDA1"),
                    yaxis = list(title = "LDA2"),
                    zaxis = list(title = "LDA3")))

subplot(p1, p2)
```


## Testing the model
```{R}
table("HDI" = test$HDI, 
      "Prediction" = predict(lda.fit, newdata = test)$class)
```
Model works very well. It correctly categorizes `r sum(test$HDI ==  predict(lda.fit, newdata = test)$class)`/`r nrow(test)` observations of the test set. 
